Tidying data

### 2025-06-13 01_tidying_data.qmd
Script updates

---
title: "Tidying data"
format: html
editor: visual
---

Setup

library(tidyverse)
library(here)
library(terra)
library(qs)

#model functions
source("src/dir.R")
source("src/model_functions.R")

#species of interest

species <- c("atlantic_salmon", 
             "general_salmonid", 
             "european_seabass", 
             "gilthead_seabream", 
             "large_yellow_croaker", 
             "japanese_seabass",
             "japanese_amberjack",
             "general_grouper",
             "pompano",
             "milkfish")

source(here("src/spatial_templates.R"))

Make repo structure

create_repo(this_path = here())

Make input and output folder structure for species of interest

#deploy data folder structure for species

map(.x = species, .f = create_species_folders)

Pull in mariculture locations data from Clawson et al 2022 and save the species groups of interest

farm_locations <- read.csv(file.path(rdsi_dir, "raw_data/aquaculture-locations/all_marine_aquaculture_farms_sources_final.csv")) |> 
  filter(species_group %in% c("Salmonidae fish", "General marine fish"))

qsave(x = farm_locations, file = "data/_general_data/farm_locations/locations.qs")

Modify Feed characterisation and save to all species forcings

#import food characterisation data - this is calculated in an independent google drive document

diet = "reference"

Food_charactisation_list <- 
  read_csv(sprintf("data/_general_data/diets/protein_lipid_carb_summaries_%s.csv", diet)) |> 
  select(-c(5)) |> 
  pivot_longer(names_to = "nutrient", values_to = "prop", cols = -species) |> 
  group_by(species) |> 
  group_split()

map(.x = Food_charactisation_list, .f = \(this_species_food){

  this_species <- this_species_food$species |> unique()

  this_species_food_df <- 
    this_species_food |> 
    relocate(prop, .before = "nutrient") |> 
    mutate(nutrient = as.character(paste0(str_to_sentence(nutrient), "s"))) |> 
    mutate(prop = as.double(prop)) |> 
    select(-species)

  write.table(this_species_food_df, file = sprintf("data/%s/forcings/Food_characterization_%s.csv", this_species, diet), row.names = FALSE, col.names = FALSE, sep = ",")

#now repeat for future feed

diet = "future"

Food_charactisation_list <- 
  read_csv(sprintf("data/_general_data/diets/protein_lipid_carb_summaries_%s.csv", diet)) |> 
  pivot_longer(names_to = "nutrient", values_to = "prop", cols = -species) |> 
  group_by(species) |> 
  group_split()

map(.x = Food_charactisation_list, .f = \(this_species_food){

  this_species <- this_species_food$species |> unique()

  this_species_food_df <- 
    this_species_food |> 
    relocate(prop, .before = "nutrient") |> 
    mutate(nutrient = as.character(paste0(str_to_sentence(nutrient), "s"))) |> 
    mutate(prop = as.double(prop)) |> 
    select(-species)

  write.table(this_species_food_df, file = sprintf("data/%s/forcings/Food_characterization_%s.csv", this_species, diet), row.names = FALSE, col.names = FALSE, sep = ",")

#now repeat for past feed

diet = "past"

Food_charactisation_list <- 
  read_csv(sprintf("data/_general_data/diets/protein_lipid_carb_summaries_%s.csv", diet)) |> 
  pivot_longer(names_to = "nutrient", values_to = "prop", cols = -species) |> 
  group_by(species) |> 
  group_split()

map(.x = Food_charactisation_list, .f = \(this_species_food){

  this_species <- this_species_food$species |> unique()

  this_species_food_df <- 
    this_species_food |> 
    relocate(prop, .before = "nutrient") |> 
    mutate(nutrient = as.character(paste0(str_to_sentence(nutrient), "s"))) |> 
    mutate(prop = as.double(prop)) |> 
    select(-species)

  write.table(this_species_food_df, file = sprintf("data/%s/forcings/Food_characterization_%s.csv", this_species, diet), row.names = FALSE, col.names = FALSE, sep = ",")

@@

nasa_dir <- "/mnt/rdsi/raw_data/NASA/GHRSST_L4_MUR_0.25_SST/"

@@

#now gapfill the saved rasters

@@

map(.x = raw_SST_files, .f = \(this_file){

  saveName <-  basename(this_file)

  message("Gapfilling ", saveName)

  this_rast <- rast(this_file)

  this_gf_rast <- focal(this_rast, w=7, fun = "mean", na.policy = "only")

  writeRaster(this_gf_rast, filename = sprintf("data/_general_data/SST/SST_gf_rasters/%s", saveName))

@@

vuln_scores <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_score.csv")) |> 
  select(vuln_gf_id, eutrophication_nutrient_pollution)

vuln_sd <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_sd.csv")) |> 
  select(vuln_gf_id, eutrophication_nutrient_pollution) |> rename(sd=eutrophication_nutrient_pollution)

@@

full_vuln_df <- 
  vuln_scores |> 
  left_join(vuln_sd) |> 
  left_join(vuln_tx)

>>>>

---
title: "Tidying data"
format: html
editor: source
---

```{r setup}
library(tidyverse)
library(here)
library(terra)
library(qs)
library(magrittr)

source("00_dirs.R")
source("00_model_functions.R")
```

```{r globals}
# Specify species of interest
species <- c(
  "atlantic_salmon",
  "general_salmonid",
  "european_seabass",
  "gilthead_seabream",
  "large_yellow_croaker",
  "japanese_seabass",
  "japanese_amberjack",
  "general_grouper",
  "pompano",
  "milkfish"
)

here("src/spatial_templates.R") %>% source()
```

# Mariculture locations

Pull in mariculture locations data from Clawson et al (2022) and save the species groups of interest.

```{r locations}
farm_locations <- file.path(rdsi_dir, "raw_data/aquaculture-locations/all_marine_aquaculture_farms_sources_final.csv") %>% 
  read.csv() %>% 
  filter(species_group %in% c("Salmonidae fish", "General marine fish"))

qsave(farm_locations, "data/_general_data/farm_locations/locations.qs")
```

@@

```{r}
nasa_dir <- "/mnt/rdsi/raw_data/NASA/GHRSST_L4_MUR_0.25_SST/"

@@

# now gapfill the saved rasters

@@

map(.x = raw_SST_files, .f = \(this_file){
  saveName <-  basename(this_file)
  message("Gapfilling ", saveName)
  this_rast <- rast(this_file)
  this_gf_rast <- focal(this_rast, w=7, fun = "mean", na.policy = "only")
  writeRaster(this_gf_rast, filename = sprintf("data/_general_data/SST/SST_gf_rasters/%s", saveName))
})
```

@@

vuln_scores <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_score.csv")) %>% 
  select(vuln_gf_id, eutrophication_nutrient_pollution)

vuln_sd <- read.csv(file.path(rdsi_data_dir, "marine_spp_vulnerabilities/butt_spp_vuln_framework-publication/_output/vuln_gapfilled_sd.csv")) %>% 
  select(vuln_gf_id, eutrophication_nutrient_pollution) %>% rename(sd=eutrophication_nutrient_pollution)

@@

full_vuln_df <- 
  vuln_scores %>% 
  left_join(vuln_sd) %>% 
  left_join(vuln_tx)
